Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2023-01-09 edubots_surveys.Rnw
Review methods - par 1,2

\begin{abstract} % abstract
Messaging platforms are applications, generally mediated by an app, desktop program or the web, mainly used for synchronous communication among users. As such, they have been widely adopted officially by higher education establishments, after little or no study of their impact and perception by the teachers. We think that the introduction of these new tools and the opportunities and challenges they have needs to be studied carefully in order to adopt the model, as well as the tool, that is the most adequate for all parties involved. We already studied the perception of these tools by students, in this paper we examine the teachers’
experiences and perceptions through a survey that we validated with peers, and what they think these tools should make or serve so that it enhances students learning and helps them achieve their learning objectives. The survey has been distributed among tertiary education teachers, both in universitary and other kind of tertiary establishments, based in
Spain (mainly) and Spanish-speaking countries. We have focused on collecting
teachers' preferences and opinions on the introduction of messaging platforms in their day-to-day work, as well as other services attached to them, such as  chatbots. What we intend with this survey is to understand their needs and to gather
information about the various educational use cases where these tools could be
valuable. In addition, an analysis of how and when teachers’ opinions towards
the use of these tools varies across gender, experience, and their discipline
of specialization is presented. The key findings of this study highlight the
factors that can contribute to the advancement of the adoption of messaging
platforms and chatbots in higher education institutions to achieve the desired
learning outcomes.
\end{abstract}

@@

New technologies are being introduced and widely adopted into the classroom in the last years. However, in order to be successful, their application
will require (many) extra work and training for the teacher, as well as for the students. Thus, both of them will acquire new skills and, potentially, pupils will
get a higher engagement to the subject, which is usually the main objective. Even if several technologies are already in use inside the classroom, there is still room for adopting others.
Anyway, this adoption strongly depends on the compromise and capabilities of the teachers, as main active actors in this new academic approach. Given this,
asking and analyzing user's opinions with regard to this scenario will help reaping the expected benefits of the technology.

In the last years, messaging applications such as WhatsApp or Telegram are being used in teaching scenarios mainly at secondary and higher education.
In addition to their main use, some of these tools offer the possibility of interact with \textit{chatbots} (or just bots), even supporting synchronous conversation between the different
parties \cite{differ.chat}. There is also possible to create a standalone chatbot application and attach it to a messaging tool through an API (Application Program Interface).

\textit{Chatbots} are autonomous agents able to communicate and interact with humans (or other chatbots) in a human-like way \cite{Gong2008}, answering question and even posing their own ones to the user.
Thus, we can say that they are close to passing a Turing Test \cite{moor2003turing}. They are implemented by means of different Artificial Intelligence techniques, such as natural language processing.
Inside the plethora of possible applications of chatbots, education is one of the most prolific nowadays \cite{clarizia2018,Smutny2020}. In this domain, they facilitate the so-called
\textit{personalized learning}, i.e. they adapt themselves to students’ pace of learning and provide online tutoring outside the classroom, which have been proved to be very effective increasing their
engagement and participation in the subjects \cite{agarwal2020}. Moreover chatbots' tutoring possibilities aid to reduce the stress of students \cite{Daniel2016,agarwal2020} by avoiding the face-to-face tutoring with their teacher,
or asking questions in front of the class. This situation normally conducts to a silence of the student and a later contact with the teacher by e-mail, which, in turn, will mean a workload for the teacher, that could be very
significant given the existing disproportion between number of students per each teacher.

@@

Given the potential benefits that chatbots could bring to the classroom, the
focus of the % EDUBOTS KA2 European project \footnote{https://www.edubots.eu},
unnamed European project, the main sponsor of this study, is to explore best practices and innovative use of chatbots,
and to create a learning community of educators in higher education institutions.
Researchers and teachers participating in this project have developed several studies. Most of these works have been focused on how the
messaging applications and/or chatbots are normally applied to deliver personalized learning in classrooms that occurs anytime anywhere,
promote collaborative learning experiences, group discussions \cite{panah2020study} and boost students' sense of belonging to their institutions
\cite{abbas2021onlinechat}.
The majority of these studies were conducted focusing on the teachers' perspectives, so this paper aims to provide more conclusions regarding
their opinions and needs, as well as the existing challenges or opportunities that arise when adopting these technologies in classrooms. This will lead to clarify
how they can positively contribute to enhancing the learning process in higher education institutions.

In this study, data collection involved two phases:
 \begin{itemize}
\item The first phase aimed to collect feedback from students about their
  students’ preferences of tools/applications for chatting and
  messaging, how they use them in educational context, who they like to be
  with in the class messaging groups, and their expectations of chatbots
  in assisting them during their learning process. Therefore, two surveys
  for bachelor and master degree students at the University of Granada (Spain)
  were designed and answers from more than 250 students were collected. The key
  findings of the students' surveys reveal that they prefer to use messaging
  familiar applications, such as Telegram or WhatsApp.
  Students also expressed interest in chatbots that can assist them with
  organizing their course schedules, help them access their assignment grades
  and facilitate searching for resources.
\item The results of the first phase were fed into the second phase of the
  study, which focused on teachers’ opinions about using chatbots and messaging
  applications in classes. These results are the main focus of this paper.
\end{itemize}

Initially, a single survey was developed for this purpose and responses from 300
higher education teachers were collected and analysed\textcolor{black}{; this will be detailed in Section \ref{sec:meth}}. The responses' analysis
led to the preparation of a second survey for teachers and 200 responses were
collected\textcolor{black}{, again following the methodology that will be explained later. The questions of the second survey focused on asking about teachers'
needs, opinions and preferences of the development of future technology-enhanced
tools and its potential impact on educational institutions policies. In order to assess these, this paper layed out the following research questions}:

@@

We will try to answer these research questions through the analysis of the two aforementioned surveys.
Eventually, what we want to obtain is a series of
recommendations to make a successful deployment of chatbot technologies (and, in
some cases, general instant messaging applications) in higher education; from the response to RQ1 and RQ4 we will try to recommend specific technologies or applications to be used in the classroom; the response to RQ2 and RQ3 will help us recommend chatbots features or specific platforms; and finally, from RQ3 we will also try to find best practices in the adoption of messaging platforms and they accompanying chatbots.

@@

% The Use of Mobile Instant Messaging (MIM) Platforms in Higher Education
The widespread and rapid adoption of free Mobile Instant Messaging
(MIM) tools/platforms such as WhatsApp, Telegram, WeChat and Facebook
Messenger stems from their simplicity, ease of use and multi-modality
(i.e. video, audio, text) \cite{tang2017mobile}. Using these tools in
higher education institutions facilitates the delivery of personalized learning
that occurs anytime anywhere, and promotes collaborative learning
experiences and group discussions \cite{panah2020study}.

WhatsApp is, at least in most Western countries, the most popular MIM platform
used by educators to provide assignments' feedback to students, support course
discussions, and contribute resources in informal learning settings
\cite{panah2020study}.  Moreover, the use of WhatsApp in higher education could
enhance social presence \cite{tang2017mobile} and foster trust relationships
between educators and students embedded in the social learning process
\cite{gachago2015crossing}. However, this last paper also reflects the need for
learners to "take ownership of the tool" and the advantages of social learning
in general. At the same time, it also mentions different challenges, among which
the most important is the blurring of social and academic life; indeed, there are challenges when using MIM tools that occur due to the
blurring of boundaries between academic and private life. This can lead to
technostress \cite{gachago2015crossing}, difficulty in managing
responsibilities, especially among mature students, and lack of privacy
\cite{tang2017mobile}.  Students' dropout of the MIM groups, as they can leave
groups at any time, can hinder their learning and undermine educators' efforts
\cite{mwakapina2016whatsapp}.  In addition, there is a need to set rules and
norms for these MIM groups in order to maintain the safety of these online
communities for students \cite{abbas2021onlinechat}. However, these rules should
not affect students' ownership and control, since it is vital to advance in
their learning \cite{gachago2015crossing} process.  This is why examining the
role of MIM platforms in higher education is still a challenge, and why the opinions of
the teaching community towards them have to be examined, as we do in this paper.

% The Use of Chatbots in Higher Education
The use of MIM platforms, although possibly valuable by itself, can be enhanced with the
use of chatbots, which are conversational agents that normally dwell 
synchronous conversations systems. The use of chatbots in higher education is still in its early stages
\cite{yang2019opportunities}. Nevertheless, recent studies have shown their
positive impact on students' academic performance \cite{perez2020rediscovering}
and engagement \cite{differ.chat,abbas2021onlinechat}, leading to a growing
interest in using this technology in the (possibly virtual) classroom. Indeed,
using chatbots to collect course feedback from students in higher
education improved response quality and boosted enjoyment levels
\cite{abbas2021Surveys}.  According to \cite{Roblyer2010}, the use of
mobile devices along with gamification strategies
\cite{Yildirim2017} can improve student motivation. In this sense, the authors
in \cite{Pimmer2019} adopted a quasi-experimental, survey-based approach to
report the positive impact of using instant messaging tools in boosting
students' knowledge and mitigating their feelings of isolation.

Several studies to evaluate chatbots in higher education have been conducted. For
instance, a comprehensive study was presented by Smutny and Schreiberova
\cite{Smutny2020} reviewing 47 educational chatbots implemented in Facebook
Messenger, with a focus on identifying characteristics and quality metrics such as
language, subject matter, and platform. Pérez and collaborators
\cite{perez2020rediscovering} attempted to categorize educational chatbots attending to their purpose 
as service-oriented or teaching-oriented. Service-oriented chatbots
include those that provide service support, such as Ask Holly
\cite{meetholly} and Dina \cite{santoso2018dinus}, which respond to
students' questions about enrollment and registration. Ask L. U.
\cite{LancasterUniversity_2019} answers students' frequently asked questions
about schedules, grades, tutors, and societies. LISA \cite{dibitonto2018chatbot}
and Differ \cite{differ.chat} facilitate breaking the ice between new students
by introducing them to each other. Ranoliya et al. \cite{ranoliya2017chatbot}
proposed a generic chatbot for university students that is able to answer 
frequently asked questions. The University of Granada deployed
Elvira, a chatbot embedded in its main web page \cite{MOREO20129} to perform the same task. In addition to being able
to answer pre-established frequent questions, it did so from the website of
the University of Granada using an inset persona, voiced by a real person, who
lip-synced the answers. However, in the case of Elvira, the emphasis was on the
authenticity of the speech-mouth and face gestures synchronization rather than making updating
answered questions simpler or more interactive for the administrative staff, or more open to the rest of the university staff using actual messaging platforms. In
most cases, the embedded search engine was able to provide more up-to-date
and accurate answers than those provided by Elvira, which had to be updated
by hand (and not too often). It was eventually discontinued, and its technology was not adapted to other platforms, since it did not really provide a useful service. Its service was not tied to teaching anyway; it was more related to directory queries and university-wide administrative questions, so even if it is strictly an example of a chatbot in a university setting, it is not really a chatbot that can be adopted by teachers to affect the learning outcomes, which is the focus of this paper.

On the other hand, teaching-oriented chatbots are more sophisticated, as they
set personalized learning outcomes and monitor learning progress. For instance,
\cite{fernoagua2018intelligent} reported on ``eduAssistant'', a virtual teaching
assistant chatbot developed on the Telegram messaging platform. In this study,
the Telegram platform was chosen because it is easy to use, students are familiar
with its features, and it enables them to exchange messages in different formats
(text, audio and video) \cite{fernoagua2018intelligent}. In addition, Telegram
can operate on all devices and operating systems. The ``eduAssistant'' chatbot
acts as an automatic agent in the teacher-content-student relationship, providing real-time
feedback loops and a personalized learning experience relevant to the
students' acquired skills and knowledge. Using this chatbot, educators can
create interactive instances in their lectures where they pose questions to
their students and the chatbot assists those who need further help by giving
them more hints and reporting it to the educator's dashboard
\cite{fernoagua2018intelligent}. This could help educators identify
students who need more attention and provide them with more educational resources
relevant to their academic achievement. %In addition, a recent study
%has analysed how the use of chatbots positively affected the learning outcomes
%of students in a Chinese class \cite{doi:10.1177/0735633120929622}.

@@

In line with this research, our previous work \cite{MoraChatbots2021:anon}
aimed at analyzing the expectations of students in this regard. Recent papers
have also analysed how the use of chatbots affected the learning outcomes of
students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper
analysed conversational chatbots in a one-to-one setting, finding that learning
truly benefited from it. This is, once again, an evidence of the benefit of
chatbots in certain settings; however, there are some prior experiences, as well as needs, that may
prevent the successful acceptance of the technology, and these are what we are
trying to find out in this paper, along with what kind of features would improve
its acceptance.

@@

\textcolor{black}{This study followed a quantitative approach towards addressing its research objective, by collecting rich data about the use of messaging applications and chatbots
by educators in both universities and colleges\footnote{In Spain, they are called simply high schools, and deliver
tertiary or associated degrees that do not have the consideration of university degrees}.
To achieve this goal two online surveys were designed and developed using Google Forms
in Spanish. They are comprised of six questions focusing on demographic data
(i.e., sector, gender, degrees, discipline, age, teaching experience)
followed by several multiple choice questions allowing the participants to choose multiple answers.}

In the {\bf first survey}, the multiple choice questions are
focused on the use of messaging apps in the teaching practice, the type of chatbot use cases that teachers consider
useful for their teaching and the impact of COVID on the teaching practice; please check them out to the full extent in the Appendix.\footnote{This {\bf first survey} was made in two separate Google forms, one for
university and the other for teachers in tertiary education institutions, outside the university system (vocational training, including 1- and 2-year degrees). Questions and
responses were the same, except for the type of tertiary degrees that
were considered.}

\textcolor{black}{This survey was piloted by the authors of the paper and their colleagues before using it in the
study.  All the comments collected from this trial were incorporated in the survey.
We mainly used mailing lists, as well as Telegram groups,
to post the survey link to reach out to more educators. The university form was sent to Spanish
(mainly Andaluc\'{\i}a and Galicia) universities, and also  universities in Costa Rica and
Mexico. The tertiary (non-university/college) teachers were based mainly in
Andaluc\'{\i}a. Dissemination and answers took place in the first quarter of
2021, during the Covid-19 pandemic and while, at least in Spain, many universities had
mandatory virtual teaching.}

A total of 282 teachers responded to the survey: 193 teaching at the University
(68.4\%) and 89 at other tertiary education institutions (31.6\%). \textcolor{black}{From those, 179 indicated their gender as male (63.5\%), 98 as female (34.8\%) and 5 teachers preferred not to indicate their gender (1.8\%). In terms of age, the majority of the participants (n=111) in this survey was 45-55 years old (39.4\%), while 91 teachers were 35-45 (32.3\%), 46 were 25-35 (16.3\%) and 34 teachers were older than 55 (12.1\%). Last, 84 teachers had a teaching experience of 16-25 years (29.8\%), 75 teachers 6-15 years of experience (26.6\%), 69 teachers had 0-5 years of experience (24.5\%) and finally, 54 teachers had more than 25 years of teaching experience (19.1\%).}

@@

% Describe the survey and population
We collected responses from 282 teachers from Spain and Spanish Speaking
countries for \textit{survey1} and 205 for \textit{survey2}. The two forms were
open for approximately the same amount of time, around two months. The sample
includes mostly teachers from university, although about 32\% and 8\% are from
non-universitary tertiary education in the first and second survey respectively. Regarding
gender, 61\% of the teachers were male and 32\% female, while approximately 5\%
chose not to disclose it. Finally, the responses are more or less equally
distributed according to the teaching experience, showing 24\% of responses from
teachers with 5 or less years of experience, 26\% for 6-15 years of experience,
30\% for 16-25, and 20\% for teachers with more than 25 years of experience.

@@

\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Kind of interactions} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5}
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
%\begin{tabular}[c]{@{}l@{}}Chat interactions among \\ students in the same course\end{tabular} & 171 & 60.6 & 110 & 39.0 \\
%\begin{tabular}[c]{@{}l@{}}Chat interactions among students, \\ teachers of the School/Faculty\end{tabular} & 13 & 4.6 & 269 & 95.4 \\
%\begin{tabular}[c]{@{}l@{}}Chat interactions among students \\ of the same study year and teachers\end{tabular} & 51 & 18.1 & 231 & 81.9 \\
%\begin{tabular}[c]{@{}l@{}}Teacher not being part of the \\ interaction\end{tabular} & 113 & 40.1 & 169 & 59.9 \\ \hline
\begin{tabular}[c]{@{}l@{}}Chat interactions among \\ students in the same course\end{tabular} & 166 & 59.1 & 115& 40.9 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students, \\ teachers of the School/Faculty\end{tabular} & 19 & 6.7 & 266 & 94.3 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students \\ of the same study year and teachers\end{tabular} & 58 & 20.5 & 224 & 79.4 \\
\begin{tabular}[c]{@{}l@{}}Teacher not being part of the \\ interaction\end{tabular} & 81 & 28.7 & 201 & 71.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Kind of interactions preferred}
\label{tab:kindinteractions}
\end{table}

@@

%A total of 205 teachers responded to the second survey, 124 males (60.5\%), 65 females (31.7\%) and 16 individuals who chose not to identify their gender (7.8\%). The majority of the teachers were aged 45-55 years old (70; 34.1\%), followed by individuals aged 35-45 (67, 32.7\%), and 25-35 (42; 20.5\%), while the minority was aged $>$~55 years old (26; 12.7\%). In terms of teaching experience, 59 teachers (28.8\%) had an experience of 16-25 years, followed by teachers who had a teaching experience of 06-15 years (51; 24.9\%), 0-5 years (51; 24.9\%) and more than 25 years of experience (44; 21.5\%). The teachers were asked about what kind of interaction media features they value the most \textcolor{black}{; i.e. what their preferred features are in the applications/tools that they use to interact with their students.} The frequencies and percentages of their responses are given in Table \ref{tab:interaction}.

@@

% No remarkable differences between different groups.
The introduction of chatbots and chatbot technology will be helped by the fact
that no discernible differences are found between different groups. Even though
chatbots do have some potential for personalization or customization based on
the class material, student, and teaching staff, the general introduction of the
technology can be done in a general way and for all disciplines, teaching
experience, and gender.

>>>>

\begin{abstract}
  Messaging platforms are applications, generally mediated by an app, desktop
  program or the web, mainly used for synchronous communication among users. As
  such, they have been widely adopted officially by higher education
  establishments, after little or no study of their impact and perception by the
  teachers. We think that the introduction of these new tools and the
  opportunities and challenges they have needs to be studied carefully in order
  to adopt the model, as well as the tool, that is the most adequate for all
  parties involved. We already studied the perception of these tools by
  students, in this paper we examine the teachers’ experiences and perceptions
  through a survey that we validated with peers, and what they think these tools
  should make or serve so that it enhances students learning and helps them
  achieve their learning objectives. The survey has been distributed among
  tertiary education teachers, both in universitary and other kind of tertiary
  establishments, based in Spain (mainly) and Spanish-speaking countries. We
  have focused on collecting teachers' preferences and opinions on the
  introduction of messaging platforms in their day-to-day work, as well as other
  services attached to them, such as chatbots. What we intend with this survey
  is to understand their needs and to gather information about the various
  educational use cases where these tools could be valuable. In addition, an
  analysis of how and when teachers’ opinions towards the use of these tools
  varies across gender, experience, and their discipline of specialization is
  presented. The key findings of this study highlight the factors that can
  contribute to the advancement of the adoption of messaging platforms and
  chatbots in higher education institutions to achieve the desired learning
  outcomes.
\end{abstract}

@@

New technologies have been introduced and widely adopted into the classroom in
recent years. However, in order to be successful, their application will require
a certain amount of extra work and training for the teacher, as well as for the
students. Training will help the two collectives involved acquire new skills
and, potentially, pupils will acquire a greater level of engagement with the
subject matter, which is usually the main objective. Even if several
technologies are already in use inside the classroom, there is still room for
adopting others.

However, the success of this adoption largely depends on the willingness and
ability of teachers to embrace it, as they are the primary agents in this new
educational approach. As such, gathering and assessing feedback from these users
will be crucial in realizing the potential benefits of the technology.

Over the past decade, messaging apps like WhatsApp and Telegram have become
increasingly popular in teaching at the secondary and higher education levels
\cite{gachago2015crossing,mwakapina2016whatsapp,yin2016adoption,panah2020study}. By
creating groups at various levels (e.g. class, assignment, course), teachers and
students have been able to use these apps for bilateral or group-level
communication. In addition to their primary function of communication, many
messaging programs also allow the creation of applications that can act as users
through open application programming interfaces (APIs), which are often
available through a free or freemium model.

This opens the door to the introduction of \textit{chatbots}, that is,
autonomous agents able to communicate and interact with humans (or other
chatbots) in a natural way \cite{Gong2008,differ.chat}, answering question and
even posing their own ones to the user, in the classroom; in this limited
environment, chatbots can appear as regular users, thus being close to passing a
Turing Test \cite{moor2003turing}, which gives them certain characteristics
that, again, can enhance student engagement. Chatbots implement different
Artificial Intelligence techniques, such as natural language processing as well
as, in some cases, learning; this is not a requirement, however, and in most
cases they are simply applications that have a conversational interface, as
opposed to a graphical user interface.

These qualities have made education one of the most promising areas for chatbot
applications today \cite{clarizia2018,Smutny2020}. In this domain, they
facilitate the so-called \textit{personalized learning}, i.e. they adapt
themselves to students’ pace of learning and provide online tutoring outside the
classroom, which have been proved to be very effective increasing their
engagement and participation in the subjects \cite{agarwal2020}. Moreover
chatbots' tutoring possibilities aid to reduce the stress of students
\cite{Daniel2016,agarwal2020} by avoiding the face-to-face tutoring with their
teacher, or asking questions in front of the class. This situation normally
conducts to a silence of the student and a later contact with the teacher by
e-mail, which, in turn, will mean a workload for the teacher, that could be very
significant given the existing disproportion between number of students per each
teacher.

@@

Given the potential benefits that chatbots could bring to the classroom, the
focus of the % EDUBOTS KA2 European project \footnote{https://www.edubots.eu},
unnamed European project, the main sponsor of this study, is to explore best
practices and innovative use of chatbots, and to create a learning community of
educators in higher education institutions.  Researchers and teachers
participating in this project have developed several studies. Most of these
works have been focused on how messaging applications and/or chatbots are
normally applied to deliver personalized learning in classrooms that occurs
anytime anywhere, and promote collaborative learning experiences or group
discussions \cite{panah2020study}, and boost students' sense of belonging to
their institutions \cite{abbas2021onlinechat}.  The majority of these studies
were conducted focusing on the teachers' perspectives, so this paper aims to
provide more conclusions regarding their opinions and needs, as well as the
existing challenges or opportunities that arise when adopting these technologies
in classrooms. This will lead to clarify how they can positively contribute to
enhancing the learning process in higher education institutions.

In this study, data collection involved two phases:
 \begin{itemize}
 \item The first phase aimed to collect feedback from students about their
   students’ preferences of tools/applications for chatting and messaging, how
   they use them in educational context, who they like to be with in the class
   messaging groups, and their expectations of chatbots in assisting them during
   their learning process. Therefore, two surveys for bachelor and master degree
   students at the University of Granada (Spain) were designed and answers from
   more than 250 students were collected. The key results from the student
   surveys indicate a preference for using familiar messaging applications, such
   as Telegram or WhatsApp. Students also expressed interest in using chatbots
   that can assist them with organizing their course schedules, help them access
   their assignment grades and facilitate searching for resources.
\item The results of the first phase were fed into the second phase of the
  study, which focused on teachers’ opinions about using chatbots and messaging
  applications in classes. These results are the main focus of this paper.
\end{itemize}

Initially, a single survey was developed for this purpose and responses from 300
higher education teachers were collected and analysed; this
  will be detailed in Section \ref{sec:meth}. The responses' analysis led to
the preparation of a second survey for teachers and 200 responses were
collected, again following the methodology that will be
  explained later. The questions of the second survey focused on asking about
  teachers' needs, opinions and preferences of the development of future
  technology-enhanced tools and its potential impact on educational institutions
  policies. In order to assess these, this paper layed out the following
  research questions:

@@

We will try to respond to these research questions through the analysis of the
answers provided by professors to the two aforementioned surveys.  Eventually,
the main research outcome of this survey should be a series of recommendations
to make a successful deployment of chatbot technologies (and, in some cases,
general instant messaging applications) in higher education; from the response
to RQ1 and RQ4 we will try to recommend specific technologies or applications to
be deployed in the classroom; the response to RQ2 and RQ3 will help us recommend
chatbots features or specific platforms; and finally, from RQ3 we will also try
to find best practices in the adoption of messaging platforms and their matching
chatbots.

@@

The widespread and rapid adoption of free Mobile Instant Messaging
(MIM) tools/platforms such as WhatsApp, Telegram, WeChat and Facebook
Messenger stems from their simplicity, ease of use and multi-modality
(i.e. video, audio, text) \cite{tang2017mobile}. Using these tools in
higher education institutions facilitates the delivery of personalized learning
that occurs anytime anywhere, and promotes collaborative learning
experiences and group discussions \cite{panah2020study}.

WhatsApp is, at least in most Western countries, the most popular MIM platform. The reason is that it is the tool preferably
used by educators to provide assignments' feedback to students, support course
discussions, and post learning resources in a semi-formal learning context
\cite{panah2020study}.  Moreover, the use of WhatsApp in higher education could
enhance social presence \cite{tang2017mobile} and foster trust relationships
between educators and students embedded in the social learning process
\cite{gachago2015crossing}. However, this last paper also reflects the need for
learners to "take ownership of the tool" and the advantages of social learning
in general. At the same time, it also mentions different challenges, among which
the most important is the blurring of social and academic life; indeed, there are challenges when using MIM tools that occur due to the
blurring of boundaries between academic and private life. This can lead to
technostress \cite{gachago2015crossing}, difficulty in managing
responsibilities, especially among mature students, and lack of privacy
\cite{tang2017mobile}.  Students' dropout of the MIM groups, as they can leave
groups at any time, can hinder their learning and undermine educators' efforts
\cite{mwakapina2016whatsapp}.  In addition, there is a need to set rules and
norms for these MIM groups in order to maintain the safety of these online
communities for students \cite{abbas2021onlinechat}. However, these rules should
not affect students' ownership and control, since it is vital to advance in
their learning \cite{gachago2015crossing} process.  This is why examining the
role of MIM platforms in higher education, as well as applications based on them such as chatbots, is still a challenge, and why the opinions of
the teaching community towards them have to be examined, as we do in this paper.

% The Use of Chatbots in Higher Education
The use of MIM platforms, although possibly valuable by itself, can be enhanced with the
use of chatbots, which are conversational agents that normally dwell in
synchronous conversations systems. The use of chatbots in higher education is still in its early stages
\cite{yang2019opportunities}. Nevertheless, recent studies have shown their
positive impact on students' academic performance \cite{perez2020rediscovering}
and engagement \cite{differ.chat,abbas2021onlinechat}, leading to a growing
interest in using this technology in the (possibly virtual) classroom. Indeed,
using chatbots to collect course feedback from students in higher
education improved the quality of responses given by students in their assignments, and boosted engagement levels
\cite{abbas2021Surveys}.  According to \cite{Roblyer2010}, the use of
mobile devices along with gamification strategies
\cite{Yildirim2017} can improve student motivation. In this sense, the authors
in \cite{Pimmer2019} adopted a quasi-experimental, survey-based approach to
report the positive impact of using instant messaging tools in boosting
students' knowledge and mitigating their feelings of isolation.

Several studies to evaluate chatbots in higher education have been conducted. For
instance, a comprehensive study was presented by Smutny and Schreiberova
\cite{Smutny2020} reviewing 47 educational chatbots implemented in Facebook
Messenger, concluding that most of these chatbots were very basic and that programmers require of more support to develop and 
offer tools for improving coaching methods and students learning. Pérez and collaborators
\cite{perez2020rediscovering} attempted to categorize educational chatbots attending to their purpose 
as service-oriented or teaching-oriented. Service-oriented chatbots
include those that provide service support, such as Ask Holly
\cite{meetholly} and Dina \cite{santoso2018dinus}, which respond to
students' questions about enrollment and registration. Ask L. U.
\cite{LancasterUniversity_2019} answers students' frequently asked questions
about schedules, grades, tutors, and societies. LISA \cite{dibitonto2018chatbot}
and Differ \cite{differ.chat} facilitate breaking the ice between new students
by introducing them to each other. Ranoliya et al. \cite{ranoliya2017chatbot}
proposed a generic chatbot for university students that is able to answer 
frequently asked questions. The University of Granada deployed
Elvira, a chatbot embedded in its main web page \cite{MOREO20129} to perform the same task. In addition to being able
to answer pre-established frequent questions, it did so from the website of
the University of Granada using an inset persona, voiced by a real person, who
lip-synced the answers. However, in the case of Elvira, the emphasis was on the
authenticity of the speech-mouth and face gestures synchronization rather than making updating
answered questions simpler or more interactive for the administrative staff, or more open to the rest of the university staff using actual messaging platforms. In
most cases, the embedded search engine was able to provide more up-to-date
and accurate answers than those provided by Elvira, which had to be updated
by hand (and not too often). It was eventually discontinued, and its technology was not adapted to other platforms, since it did not really provide a useful service. Its service was not tied to teaching anyway; it was more related to directory queries and university-wide administrative questions, so even if it is strictly an example of a chatbot in a university setting, it is not really a chatbot that can be adopted by teachers to affect the learning outcomes, which is the focus of this paper.

Contrarily to the mentioned chatbots that are focused on administrative tasks, teaching-oriented chatbots are more sophisticated, as they
set personalized learning outcomes and monitor learning progress. For instance,
\cite{fernoagua2018intelligent} reported on ``eduAssistant'', a virtual teaching
assistant chatbot developed on the Telegram messaging platform. In this study,
the Telegram platform was chosen because it is easy to use, students are familiar
with its features, and it enables them to exchange messages in different formats
(text, audio and video) \cite{fernoagua2018intelligent}. In addition, Telegram
can operate on all devices and operating systems. The ``eduAssistant'' chatbot
acts as an automatic agent in the teacher-content-student relationship, providing real-time
feedback loops and a personalized learning experience relevant to the
students' acquired skills and knowledge. Using this chatbot, educators can
create interactive instances in their lectures where they pose questions to
their students and the chatbot assists those who need further help by giving
them more hints and reporting it to the educator's dashboard
\cite{fernoagua2018intelligent}. This could help educators identify
students who need more attention and provide them with more educational resources
relevant to their academic achievement. %In addition, a recent study
%has analysed how the use of chatbots positively affected the learning outcomes
%of students in a Chinese class \cite{doi:10.1177/0735633120929622}.

@@

In line with this research, our previous work \cite{MoraChatbots2021:anon}
aimed at analyzing the expectations of students in this regard. Another paper
have also analysed how the use of chatbots affected the learning outcomes of
students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper
considered conversational chatbots in a one-to-one setting, finding that learning
truly benefited from it. This is, once again, an evidence of the benefit of
chatbots in certain settings; however, there are some prior experiences, as well as needs, that may
prevent the successful acceptance of the technology, and these are what we are
trying to find out in this paper, along with what kind of features would improve
its acceptance.

@@

This study has employed a quantitative approach in order to address its research objectives. A good amount of data was collected on the usage of messaging applications and chatbots by educators in universities and colleges.\footnote{In Spain, higher education institutions generally known elsewhere as colleges, are called simply high schools, and deliver
tertiary or associate degrees that do not have the consideration of university degrees.}
To achieve this goal two online surveys were designed and developed using Google Forms
in Spanish. They are comprised of six questions focusing on demographic data
(i.e., sector, gender, degrees, discipline, age, or teaching experience)
followed by several multiple choice questions allowing the participants to choose multiple answers.

In the {\bf first survey}, multiple choice questions were
focused on the use of messaging apps in the teaching practice, the type of chatbot use cases that educators would consider
useful for their teaching and the impact of COVID on the teaching practice; please, check them out to the full extent in the Appendix.\footnote{This {\bf first survey} was made in two separate Google forms, one for
university and another for teachers in tertiary education institutions, outside the university system (vocational training, including 1- and 2-year degrees). Questions and
responses were the same, except for the type of tertiary degrees that
were considered.} This survey was piloted by the authors of this paper and their colleagues before using it in the
study. All the comments collected from this trial were incorporated in the survey.
We primarily used mailing lists and Telegram groups to disseminate the survey link so that it would reach a larger number of educators. The form targeting university teachers was sent to Spanish
(mainly Andaluc\'{\i}a and Galicia) universities, and also universities in Costa Rica and
Mexico. The tertiary (non-university/college) educators who received the form were based mainly in
Andaluc\'{\i}a. Dissemination and answers took place in the first quarter of
2021, during the Covid-19 pandemic and while, at least in Spain, many universities had
mandatory virtual teaching.

A total of 282 teachers responded to the survey: 193 teaching at the University
(68.4\%) and 89 at other tertiary education institutions (31.6\%). From those, 179 indicated their gender as male (63.5\%), 98 as female (34.8\%) and 5 teachers preferred not to indicate their gender (1.8\%). In terms of age, the majority of the participants (n=111) in this survey was 45-55 years old (39.4\%), while 91 teachers were 35-45 (32.3\%), 46 were 25-35 (16.3\%) and 34 teachers were older than 55 (12.1\%). Last, 84 teachers had a teaching experience of 16-25 years (29.8\%), 75 teachers 6-15 years of experience (26.6\%), 69 teachers had 0-5 years of experience (24.5\%) and finally, 54 teachers had more than 25 years of teaching experience (19.1\%).

@@

% Describe the survey and population
We collected responses from 282 teachers from Spain and Spanish-speaking
countries for \textit{survey1} and 205 for \textit{survey2}. The two forms were
open for approximately the same amount of time, around two months. The sample
includes mostly teachers from university, although about 32\% and 8\% are from
non-universitary tertiary education in the first and second survey respectively. Regarding
gender, 61\% of the teachers were male and 32\% female, while approximately 5\%
chose not to disclose it. Finally, the responses are more or less equally
distributed according to the teaching experience, showing 24\% of responses from
teachers with 5 or less years of experience, 26\% for 6-15 years of experience,
30\% for 16-25, and 20\% for teachers with more than 25 years of experience.

@@

\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Kind of interactions} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5}
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
\begin{tabular}[c]{@{}l@{}}Chat interactions among \\ students in the same course\end{tabular} & 166 & 59.1 & 115& 40.9 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students, \\ teachers of the School/Faculty\end{tabular} & 19 & 6.7 & 266 & 94.3 \\
\begin{tabular}[c]{@{}l@{}}Chat interactions among students \\ of the same study year and teachers\end{tabular} & 58 & 20.5 & 224 & 79.4 \\
\begin{tabular}[c]{@{}l@{}}Teacher not being part of the \\ interaction\end{tabular} & 81 & 28.7 & 201 & 71.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Kind of interactions preferred}
\label{tab:kindinteractions}
\end{table}

@@

@@

The introduction of chatbots and chatbot technology will be helped by the fact
that no discernible differences are found between different groups. Even though
chatbots do have some potential for personalization or customization based on
the class material, student, and teaching staff, the general introduction of the
technology can be done in a general way and for all disciplines, teaching
experience, and gender.
